Higher-Order unification for free!

### 2024-04-15 paper/main.tex
6 done, some todo

\subsection{Substitution decompilation}

@@

In \cref{sec:grounwork}, we introduced the definition of ``memory map'' (\mapStore).
This memory allows to decompile the \Ho terms back to the object language.
It is the case that, while solving unification problems, a same unification
variable $X$ is used multiple times with different arities.
%
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{X,E,1},
    {Y,F,0},
    {X,C,2}}}
  {{{\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {\eta,,A,\lambda x.B_{x}},
    {\eta,x,B_{x},\lambda y.C_{y x}}}}

In the unification problems \foUnifPb above, we see that $X$ is used with arity $2$ in
$\foUnifPb_1$ and with arity $1$ in $\foUnifPb_2$. 
By \fullref{inv:uvaarity}, we are not allowed to use a same \Ho variable to 
represent the two occurrences of $X$.
If we execute \hrun, we remark that the unification fails.
There is in fact a major problem:
\hstep is not conscious of the connection between the variables $C$ and $E$ (both corresponding to $X$),
since no link in \linkStore puts $C$ and $E$ in relation and decompilation
does not work properly if a \Fo{} variable is mapped to two distinct \Ho variables. 
The two main drawbacks connected to this situation are 
firstly the lost of \fullref{prop:fidelity} and
secondly, if we want to guarantee at least \fullref{prop:simulation}, we should
overcomplicate the decompilation phase. In order to ease the second
drawback, we pose the following property:

\begin{proposition}[\mapStore is a bijection]
  Given a list of unification problems \foUnifPb, then the memory map \mapStore
  compiled from \foUnifPb is a bijection relating the \Fo{} and the \Ho{} 
  variables.
\end{proposition}

We finally adjust the compiler's output with a \elpiIn{map-deduplication} procedure.

@@

\begin{theorem}[Fidelity with \mapdeduplication]
  Given a list of unification problems \foUnifPb, such that 
  $\forall t, t\in\subterm{\foUnifPb} \Rightarrow \wellb(t) \lor t \in \maybeeta$,
  if \foUnifPb contains two same \Fo{} variables with different arities,
  then \mapdeduplication guarantees \fullref{prop:fidelity}
\end{theorem}

\begin{proof}[Proof sketch]
  By the definition of \mapdeduplication, any two occurrencies of the same \Fo{} 
  variables $X_1, X_2$
  with different arities are related with \linketa. 
  If one of the two variables is instantiated, the corresponding
  \linketa is triggered instantiating the related variable.
  This allows to make unification fail if $X_1$ and $X_2$ are unified with 
  different terms. Finally, since \foUnifPb contains only terms thar are
  either \wellb or \maybeeta, by \cref{lemma:fidelity-maybeeta},
  we can conclude the proof.
\end{proof}

@@

In this example, $\hoUnifPb_1$ assigns $A$ which triggers $\linkStore_3$ and
then $\linkStore_4$ by \progressetaleft. $C_{yx}$ is therefore
assigned to $x$ (the second variable of its scope). We can finally see the
\progressetaright of $\linkStore_1$: its \rhs is now $\lambda y.y$ (the term $C_{xy}$
reduces to $y$). Since it is no more in \maybeeta, $\lambda y.y$ is unified with
$E_x$. After the execution of the remaining \hstep, we obtain the 
following \Fo{} substitution $\rho = \{X := \lambda x.\lambda y.y, Y := (f\appsep \lambda x.x)\}$.

>>>>

\subsection{Substitution decompilation}\label{sec:decompilation}

@@

In \cref{sec:grounwork}, we introduced the definition of memory map \mapStore.
This map allows to decompile the \Ho substitution to the object language.
Given the higher order nature of CIC the same unification
variable $X$ is used multiple times with different arities.
%
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{X,E,1},
    {Y,F,0},
    {X,C,2}}}
  {{{\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {\eta,,A,\lambda x.B_{x}},
    {\eta,x,B_{x},\lambda y.C_{y x}}}}

In the unification problems \foUnifPb above, we see that $X$ is used with arity $2$ in
$\foUnifPb_1$ and with arity $1$ in $\foUnifPb_2$. 
In order to preserve \fullref{inv:uvaarity} the compiler
did generate two entries in \mapStore for $X$, and in turn this
breaks \fullref{prop:fidelity} since two incompatible assignements to
the variables $X$ maps to, namely $E$ and $C$, would not be detected.
We remedy by post-processing the compiler's output by adding
additional \linketa to ensure the folllowgin property:

\begin{proposition}[\mapStore is a bijection]
  After compilation, for each \Fo{} variable $X$ in \foUnifPb{} and for each \Ho{}
  variable $A$ in \hoUnifPb{} there is exactly one entry $X \mapsto A^n$
  (for any arity $n$).
\end{proposition}

\noindent
Note that the execution in \Ho{} may require allocating new variables as
explained in~\cref{sec:decompilation}. The property above does not
concern these.

The core procedure of this post-processing step is \emph{align-arity} that
is iterated by \elpiIn{map-deduplication}:

@@

\begin{theorem}[Fidelity with \mapdeduplication]
  Given a list of unification problems \foUnifPb, such that 
  $\forall t, t\in\subterm{\foUnifPb} \Rightarrow \wellb(t) \lor t \in \maybeeta$,
  if $\;\foUnifPb$ contains the same \Fo{} variable used at different arities,
  then \mapdeduplication guarantees \fullref{prop:fidelity}
\end{theorem}

\begin{proof}[Proof sketch]
  By the definition of \mapdeduplication, any two occurrencies of the same \Fo{} 
  variables $X_1, X_2$ with different arities are compiled to two
  variables $A^n$ and $C^m$ related by a chain of \linketa.
  We show that whenever one of the two is instantiated with a rigid term the 
  the chain becomes shorter and hence, eventually, the
  other variable is instantiated too.
  Without loss of generality we assume $m < n$ and $A$ is instantiated.
  The corresponding \linketa is triggered by \progressetaleft instantiating
  the next variable in the chain of links.
\end{proof}
\todo{Not sure I fullfy fixed the proof, but the old oen was not ok}

\noindent
Note that shortening the chain of links allows unification to fail if
$X_1$ and $X_2$ are unified with different terms.

@@

\noindent
The resolution of $\hoUnifPb_1$ assigns $\lambda x.\lambda y.x$ to $A$
that in turn triggers $\linkStore_3$ and then $\linkStore_4$ by \progressetaleft,
one per $\lambda$ in the solution of $A$.
The unification variable $C_{yx}$ is therefore unified with $x$
(the second variable of its scope).
As a result $\linkStore_1$ becomes \linketaM{x}{E_x}{\lambda y.y},
and the \rhs is no more in \maybeeta so \progressetaright fires and
$E_x$ is unified with $\lambda y.y$, resulting in
$\sigma = \{~ A \mapsto \lambda x.\lambda y.x ~;~ B_x \mapsto \lambda y.x ~;~ C_{yx} \mapsto x ~;~ E_x \mapsto \lambda y.y~\}$.\\
The remaining simplify step $\hoUnifPb_2$ identified $D$ with $F$,
hence the resulting \Fo{} substitution is
$\rho = \{X \mapsto \lambda x.\lambda y.y ~;~ Y \mapsto \lambda x.(f\appsep (\lambda y.y))\appsep x\}$.
\todo{verificare, la vecchia soluzione era sballata}
